\section{Assessing the Motivations Relating to Correlation Between Workers}

\subsection{correlation Between Workers}
Empirical evidence of cross‑worker and temporal correlations in gradients
A key question is whether the right form of correlation exists for WZ decoding to be effective. Our decoder aims to reconstruct an accumulated gradient for the next update using a short history of past reconstructions (from the same worker and, in the all‑workers variant, from other workers as well). Measuring correlation at the wrong granularity can be misleading: comparing a single per‑batch gradient across workers tends to show weak correlation (small steps on small data subsets), while comparing only one accumulated gradient per round yields too few pairs and mixes signals that are many epochs apart which doesn't tell us how well neighboring rounds are related. We therefore evaluate accumulated gradients within a round—a sliding window over the last  steps—so we obtain many pairs at meaningful temporal distances. Empirically, this reveals consistent cross‑worker similarity at the scale relevant to reconstruction (figure to be added).


\subsection{Distribution Shift}
Finally, we account for distribution shift. WZ decoders are sensitive to changes in the distribution of inputs and targets. In our setting, the server learns to reconstruct from past reconstructions, so drift over rounds can compound error. Two design choices address this: (i) per‑layer normalization (e.g., 99th‑percentile scaling) keeps dynamic ranges comparable across layers and rounds; and (ii) a hybrid switching policy that transitions from all‑workers DSC to a worker‑only regime after a fixed number of rounds, once the MSPE begins to rise. Together, these choices preserve the benefits of side information early while maintaining stability later.