% Chapter 3: Materials and Methods
\section{The Gains from Wyner-Ziv Quantization and rANS Coding}

% • Symbol probability modeling and entropy coding
% • Realized‑bits reporting (gzip in prototype; rANS‑compatible design)

The compression gains of the WZ quantizer are split into three parts.
The first part is not DSC, and is related to the fact that quantizing a floating-point number usually yields a number with a smaller bit size, if done correctly. This gain is not separately visible when using a DSC coder, as the DSC coder will improve on this gain.
The second is the direct gain of the DSC quantizer and is because we are having larger, fewer bins, and reusing bins in different regions of the input space. We call this the marginal gain and it is the same as the entropy of the marginal prior from \cite{OzyilkanRecoverBinning}. This gain is directly visible when using a DSC coder and would be true per single input value.
The third gain is more indirect and requires extra steps as it relates to compressing a vector of the quantized values. This gain is related to the conditional prior from \cite{OzyilkanRecoverBinning} and is because a long list of quantized values could be compressed better when conditioned on the side information, hence the name conditional gain. To be able to leverage this gain, we require a DSC entropy coder which is aware of the conditional prior. But this conditional prior is usually only available on one side, so it requires a separate DSC coding which incorporates the prior in its coding scheme. We did not want to get involved with the design of a new DSC coder, but to be able to see the realized gains of our WZ quantizer, we opt to use a non-DSC entropy coder alongside the prior to simulate the effect of a DSC coder.

rANS, or range Asymmetric Numeral Systems, is a modern entropy coding method that provides efficient compression by encoding data based on symbol probabilities. It is particularly well-suited for scenarios where the probability distribution of symbols conditioned on side information (the conditional prior) is known or can be estimated, making it ideal for our Wyner-Ziv quantization scheme.